---
title: "Class 9: Candy Mini-Project"
author: "Aadhya Tripathi (PID: A17878439)"
toc: true
format: pdf
---

## Background

In today's mini-project we will analyze candy data with exploratory graphics, ggplot, basic statistics, correlation analysis, and principal component analysis methods we have been learning.

## Data Import

The data comes as a CSV file from FiveThirtyEight.

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` different candy types.

> Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` fruity candy types.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is "Milky Way" with a winpercent of `r candy["Milky Way", ]$winpercent`.

> Q4. What is the winpercent value for “Kit Kat”?

The winpercent for "Kit Kat" is `r candy["Kit Kat", ]$winpercent`.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The winpercent for "Tootsie Roll Snack Bars" is `r candy["Tootsie Roll Snack Bars", ]$winpercent`.

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```
The winpercent variable is on a different scale in comparison to the other columns.

>  Q7. What do you think a zero and one represent for the candy$chocolate column?

It is a true/false for whether the candy type is chocolate or not.

## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

Using base R:
```{r}
hist(candy$winpercent)
```
Using ggplot:
```{r}
library(ggplot2)
```

```{r}
ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(bins = 10, fill = "lightblue", col = "darkblue")
```

> Q9. Is the distribution of winpercent values symmetrical?

No, there is a skew to the right.

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
summary(candy$winpercent)
```
The mean is slightly above 50%. However, the median is below 50%. Based on the median, the center is determined to be below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Steps to solve this problem:
1. Find all chocolate candy in the dataset
2. Find winpercent of these
3. Calculate mean winpercent
4. Repeat 1-3 for fruity candy
5. Compare chocolate mean and fruity mean

```{r}
choc_win <- candy$winpercent[as.logical(candy$chocolate)]
fruity_win <- candy$winpercent[as.logical(candy$fruity)]

print(mean(choc_win))
print(mean(fruity_win))
mean(choc_win) > mean(fruity_win)
```
Based on their mean values, chocolate candy is ranked higher than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc_win, fruity_win)
```
Based on the extremely low p-value, we can reject the null hypothesis and the difference is statistically significant.

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

Use dplyr to rearrange the data by winpercent. 

```{r}
library(dplyr)
```

```{r}
candy |>
  arrange(winpercent) |>
    head(5)
```

The 5 least liked candy types in this dataset are "Nik L Nip", "Boston Baked Beans", "Chiclets", "Super Bubble", and "Jawbusters".

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(desc(winpercent)) |>
    head(5)
```

The 5 all time favorite candy types in this dataset are "Reese's Peanut Butter cup", "Reese's Miniatures", "Twix", "Kit Kat", and "Snickers".

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=10, fig.width=6}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent.

```{r, fig.height=10, fig.width=6}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() +
  ylab("")
```

Color the bars based on candy type. Brown is for chocolate, blue for bar candies, and green for fruity candies.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "#BD5C00"
my_cols[as.logical(candy$bar)] = "#458CED"
my_cols[as.logical(candy$fruity)] = "#0BAD00"
```

```{r, fig.height=10, fig.width=6}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```
> Q17. What is the worst ranked chocolate candy?

Sixlets.

> Q18. What is the best ranked fruity candy?

Starburst.

## Taking a look at pricepercent

Use ggrepel to make overalpping data point labels easier to read. 
```{r}
library(ggrepel)
```

Make a plot of winpercent vs pricepercent:
```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures have a high winpercent while having a relatively low pricepercent. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy |>
  arrange(desc(pricepercent)) |>
    head(5)
```

The top 5 most expensive candy types are "Nik L Nip", "Nestle Smarties", "Ring pop", "Hershey's Krackel", and "Hershey's Milk Chocolate". The least popular of these is "Nik L Nip".

## Exploring the correlation structure

Pearson correlation values range from -1 to +1. 

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

The main results figure is the PCA score plot:

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols) +
        labs(title="PCA Candy Space Map")

p
```

```{r}
p <- p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (brown), bar (blue), fruity (green), other (black)",
       caption="Data from 538")

p
```
Make an interactive plot with plotly (excluded for PDF render):

```{r}
# library(plotly)
```

```{r}
# ggplotly(p)
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col() +
  ylab("")
```

Fruity, pluribus, and hard are in the positive direction. It makes sense for these characteristics to be correlated as many fruity candies come in multiples and may be hard more often than chocolate. This relationship was previously highlighted in the correlation matrix.

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Chocolate, bar candy types appear to be the "winning" candy. The correlation matrix shows that winpercent has strongest positive correlation with chocolate, and it has a medium positive correlation with bar as well. The PCA plot supports that chocolate and bar candy are more similar to each other compared to fruity candy, as they are closer together on the PC1 axis than chocolate and fruity. The winpercent ordered barplot visualization shows that that brown and blue, representing chocolate and bar candies, tend to be higher compared to the green fruity candies.

