---
title: "Class 12, pt. 2: Population Analysis"
author: "Aadhya Tripathi (PID: A17878439)"
toc: true
format: pdf
---

## Population Scale Analysis

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
# Import the file
results <- read.table("rs8067378_ENSG00000172057.6.txt")
head(results)
```

Gives the sample size of each genotype:
```{r}
table(results$geno)
```
Calculate the median expression:
```{r}
AA <- results[results$geno == "A/A",]
AG <- results[results$geno == "A/G",]
GG <- results[results$geno == "G/G",]

median(AA$exp)
median(AG$exp)
median(GG$exp)
```
A|A has a sample size of 108 and median expression 31.24847.
A|G has a sample size of 233 and median expression 25.06486
G|G has a sample size of 121 and median expression 20.07363

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

Boxplot of the genotype vs expression:
```{r}
library(ggplot2)
```

```{r}
boxp <- ggplot(results) +
        aes(geno, exp, fill = geno) +
        geom_boxplot()

boxp
```

The G/G genotype decreases expression compared to A/A, based on the lower median value for G/G. The SNP decreases the expression of ORMDL3.
