---
title: "Class 17: AWS Analysis"
author: "Aadhya Tripathi (PID: A17878439)"
toc: true
format: pdf
---

## Downstream Analysis of Kallisto Output

```{r load tximport}
library(tximport)
```

```{r import kallisto files}
# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

Number of transcripts per sample:
```{r}
colSums(txi.kallisto$counts)
```

Number of transcripts detected in at least one sample:
```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filter out annotated transcripts with no reads:
```{r filter no reads}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

Filter out transcripts with no change over the samples:
```{r filter no change}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

Run PCA, scaled based on transcipt levels:
```{r run pca}
pca <- prcomp(t(x), scale=TRUE)
```

```{r pca summary}
summary(pca)
```

Plot of first 2 principal components:
```{r basic plot of PC1 vs PC2}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3

```{r load ggplot}
library(ggplot2)
library(ggrepel)
```

```{r}
mycols <- c("blue","blue","red","red")
```

```{r ggplot PC1 vs PC2}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw() +
  labs(title = "PC1 vs PC2")
```

```{r ggplot PC1 vs PC3}
ggplot(pca$x) +
  aes(PC1, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw() +
  labs(title = "PC1 vs PC3")
```

```{r ggplot PC2 vs PC3}
ggplot(pca$x) +
  aes(PC2, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw() +
  labs(title = "PC2 vs PC3")
```